---
title: "Final Project"
format: html
editor: visual
---

## Set Global Seed

```{r setup, include=FALSE}
set.seed(1969)
```

## Data Generation

First test set at inoc size 10 and sd 0.2

Growth parameter values generally assumed

```{r}
nutrient_levels <- c(0.25,0.5,1,1.5,2.5,5,7.5,10,20,100,200,500)

# Parameters for growth model
mumax <-1
Qmin <-1
inoc<-1
sd_variation <- 0.02

growth_data <- data.frame()

for (R in nutrient_levels) {
  # Calculate mean growth rate for this nutrient level
  mean_growth <- mumax * R / (inoc*Qmin+ R)
  
  # Generate 6 replicate growth rates
  replicates <- rnorm(6, mean = mean_growth, sd = sd_variation)
  # replicates <- pmax(replicates, 0.0001)  # ensure non-negative
  
  growth_data <- rbind(growth_data, data.frame(
    nutrient = R,
    growth_rate = replicates,
    replicate = 1:6
  ))
}

print(growth_data)

ggplot(growth_data, aes(x = nutrient, y = growth_rate)) +
  geom_point() +
  labs(x = "Nutrient Level", y = "Growth Rate") +
  theme_classic()

```

It works!

↓ Celebratory Claps ↓

```{r}
#install.packages("tuneR")
library(tuneR)
my_audio <- readMP3("applause-sound-427084.mp3")
tuneR::setWavPlayer('/usr/bin/aplay')
play(my_audio)
```

## Fitting Monod Curve to Data

first test model

```{r}
# Intercept only for both paramters as we only want one estimation per value for each dataset defined by the inoculum size and sd
monod_formula <- bf(
  growth_rate ~ mumax * nutrient / (K + nutrient),
  mumax ~ 1,
  K ~ 1,     
  nl = TRUE    # non-linear model
)

# Set Priors
priors <- c(
  prior(normal(1, 0.5), nlpar = "mumax", lb = 0), 
  prior(normal(0.0101, 1000), nlpar = "K", lb = 0)       
)

# Fit Bayesian Model
Test_Model<-brm(formula=monod_formula,
                data = growth_data,
                prior = priors,
                family =gaussian(),
                chains = 4,
                iter = 4500,
                warmup = 2000,
                cores = 6,
                control = list(adapt_delta = 0.95)
)

# View results
summary(Test_Model)

# Extract parameter estimates
posterior_samples <- as_draws_df(Test_Model)
mumax_median <- median(posterior_samples$b_mumax_Intercept)
K_median <- median(posterior_samples$b_K_Intercept)
mumax_mean <- mean(posterior_samples$b_mumax_Intercept)
K_mean <- mean(posterior_samples$b_K_Intercept)

# Plot posterior distributions
plot(Test_Model, variable = c("b_mumax_Intercept", "b_K_Intercept"))

# Create prediction plot with credible intervals
new_nutrient <- data.frame(nutrient = seq(0, 500, length.out = 500))
predictions <- predict((Test_Model), newdata = new_nutrient)

pred_data <- cbind(new_nutrient, predictions)

ggplot() +
  geom_point(data = growth_data, aes(x = nutrient, y = growth_rate), 
             size = 2, alpha = 0.6) +
  geom_line(data = pred_data, aes(x = nutrient, y = Estimate), 
            color = "blue", size = 1) +
  geom_ribbon(data = pred_data, 
              aes(x = nutrient, ymin = Q2.5, ymax = Q97.5), 
              alpha = 0.2, fill = "blue") +
  labs(title = "Bayesian Monod Curve Fit",
       x = "Nutrient Concentration",
       y = "Growth Rate",
       subtitle = "Blue line = median prediction, shaded area = 95% credible interval") +
  theme_classic()

# Check model fit
pp_check(Test_Model, ndraws = 100)

# Compare prior vs posterior (optional)
prior_summary(Test_Model)

print(Test_Model)

mcmc_pairs(Test_Model, pars = c("b_mumax_Intercept", "b_K_Intercept", "sigma"), off_diag_args = list(size = 1.5))
plot(Test_Model)
posterior_interval(Test_Model)
```

It works!

## Analytic Solution

```{r}
find_growth_rate <- function(mumax,Vmax,R,H,Qmin) {
  
  growth_rate <- (R*Vmax*mumax)/(H*Qmin*mumax+(R*(Vmax+(Qmin*mumax))))
  
  return(growth_rate)
}

find_k <- function(mumax,Vmax,R,H,Qmin) {
  
  k<-((mumax*Qmin*H)/(Vmax-(mumax*Qmin)))
  
  return(k)
}


find_growth_rate(mumax=1,Vmax=100,R=0.0101,H=1,Qmin=1)
find_k(mumax=1,Vmax=100,R=1,H=1,Qmin=1)

# Param values of our theoretical microbe
mumax=1
Vmax=100
H=1
Qmin=1

analytic_function <- function(R) {
  (R*Vmax*mumax)/(H*Qmin*mumax+(R*(Vmax+(Qmin*mumax))))
}

ggplot()+
  geom_point() +
  labs(x="Nutrient Level",y="Growth Rate")+
  stat_function(fun = analytic_function)+
  theme_classic()
```

This is what we will compare model fits to

## check model fit between stochastic families

```{r}
nutrient_levels <- c(0.25,0.5,1,1.5,2.5,5,7.5,10,20,100,200,500)

# Parameters for growth model
mumax <-1
Qmin <-1
inoc<-1
sd_variation <- 0.02

growth_data <- data.frame()

for (R in nutrient_levels) {
  # Calculate mean growth rate for this nutrient level
  mean_growth <- mumax * R / (inoc*Qmin+ R)
  
  # Generate 6 replicate growth rates
  replicates <- rnorm(6, mean = mean_growth, sd = sd_variation)
  replicates <- pmax(replicates, 0.0001)  # ensure non-negative
  
  growth_data <- rbind(growth_data, data.frame(
    nutrient = R,
    growth_rate = replicates,
    replicate = 1:6
  ))
}

print(growth_data)

ggplot(growth_data, aes(x = nutrient, y = growth_rate)) +
  geom_point() +
  labs(x = "Nutrient Level", y = "Growth Rate") +
  theme_classic()

monod_formula <- bf(
  growth_rate ~ mumax * nutrient / (K + nutrient),
  mumax ~ 1,
  K ~ 1,     
  nl = TRUE    # non-linear model
)

# Set Priors
priors <- c(
  prior(normal(1, 0.5), nlpar = "mumax", lb = 0), 
  prior(normal(0.0101, 3000), nlpar = "K", lb = 0)       
)

# Fit Bayesian Model
Test_Model<-brm(formula=monod_formula,
                data = growth_data,
                prior = priors,
                family =gaussian(),
                chains = 4,
                iter = 4500,
                warmup = 2000,
                cores = 6,
                control = list(adapt_delta = 0.95)
)

# View results
summary(Test_Model)

# Extract parameter estimates
posterior_samples <- as_draws_df(Test_Model)
mumax_median <- median(posterior_samples$b_mumax_Intercept)
K_median <- median(posterior_samples$b_K_Intercept)
mumax_mean <- mean(posterior_samples$b_mumax_Intercept)
K_mean <- mean(posterior_samples$b_K_Intercept)

# Plot posterior distributions
plot(Test_Model, variable = c("b_mumax_Intercept", "b_K_Intercept"))

# Create prediction plot with credible intervals
new_nutrient <- data.frame(nutrient = seq(0, 500, length.out = 500))
predictions <- predict((Test_Model), newdata = new_nutrient)

pred_data <- cbind(new_nutrient, predictions)

ggplot() +
  geom_point(data = growth_data, aes(x = nutrient, y = growth_rate), 
             size = 2, alpha = 0.6) +
  geom_line(data = pred_data, aes(x = nutrient, y = Estimate), 
            color = "blue", size = 1) +
  geom_ribbon(data = pred_data, 
              aes(x = nutrient, ymin = Q2.5, ymax = Q97.5), 
              alpha = 0.2, fill = "blue") +
  labs(title = "Bayesian Monod Curve Fit",
       x = "Nutrient Concentration",
       y = "Growth Rate",
       subtitle = "Blue line = median prediction, shaded area = 95% credible interval") +
  theme_classic()

# Check model fit
pp_check(Test_Model, ndraws = 100)

# Compare prior vs posterior
prior_summary(Test_Model)

print(Test_Model)

summary(Test_Model)

mcmc_pairs(Test_Model, pars = c("b_mumax_Intercept", "b_K_Intercept", "sigma"), off_diag_args = list(size = 1.5))

plot(Test_Model)

Test_Model_Gamma_log<-brm(formula=monod_formula,
                data = growth_data,
                prior = priors,
                family =Gamma(link="log"),
                chains = 4,
                iter = 4500,
                warmup = 2000,
                cores = 6,
                control = list(adapt_delta = 0.95)
)

Test_Model_Gamma_identity<-brm(formula=monod_formula,
                          data = growth_data,
                          prior = priors,
                          family =Gamma(link="identity"),
                          chains = 4,
                          iter = 4500,
                          warmup = 2000,
                          cores = 6,
                          control = list(adapt_delta = 0.95)
)


LOO(Test_Model,Test_Model_Gamma_log,Test_Model_Gamma_identity)
```

## Code to run all simulation and fit

Building in the different inoc and variances of the simulated data

```{r}
# Fitting Monod models across multiple inoculum x SD combinations
library(brms)
library(tidyverse)

# Define parameter combinations
inoculum_sizes <- c(0.005,0.01,0.02,0.025,0.05,0.1,0.15,0.25,0.5,1,5,10)  # 12 levels
std_devs <- c(0,0.05,0.1,0.15,0.2,0.25,0.3)        # 7 levels

# Create all combinations (5 x 5 = 25)
param_grid <- expand.grid(
  inoculum = inoculum_sizes,
  sd = std_devs
)

# Define Monod formula (same for all fits)
monod_formula <- bf(
  growth_rate ~ mumax * nutrient / (K + nutrient),
  mumax ~ 1,
  K ~ 1,     
  nl = TRUE    # non-linear model
)

# Set Priors
priors <- c(
  prior(normal(1,1), nlpar = "mumax", lb = 0),
  prior(normal(0.0101, 3000), nlpar = "K", lb = 0)
)

# Initialize results storage
results <- data.frame(
  inoculum = numeric(),
  sd = numeric(),
  mumax_estimate =numeric(),
  mumax_estimate_error = numeric(),
  mumax_lower = numeric(),
  mumax_upper = numeric(),
  mumax_bulk_ess=numeric(),
  mumax_tail_ess=numeric(),
  mumax_rhat =numeric(),
  K_estimate = numeric(),
  K_estimate_error = numeric(),
  K_lower =numeric(),
  K_upper = numeric(),
  K_bulk_ess=numeric(),
  K_tail_ess=numeric(),
  K_rhat =numeric(),
  sigma=numeric()
)
```

```{r}
# Loop through all combinations
set.seed(1969)
nutrient_levels <- c(0.001,0.005,0.01,0.05,0.1,0.5,1,5)
start_time_total<-Sys.time()
for (i in 1:nrow(param_grid)) {
  
  cat("\n=== Fitting model", i, "of", nrow(param_grid), "===\n")
  cat("Inoculum:", param_grid$inoculum[i], "| SD:", param_grid$sd[i], "\n")
  start_time<-Sys.time()
  
  current_inoculum <- param_grid$inoculum[i]
  current_sd <- param_grid$sd[i]
  
  mumax <- 1
  Qmin <- 1
  
  growth_data <- data.frame()
  for (nutrient in nutrient_levels) {
    mean_growth <- mumax * nutrient / (current_inoculum*Qmin+ nutrient)
    replicates <- rnorm(6, mean = mean_growth, sd = current_sd)
    replicates <- pmax(replicates, 0.001)  # ensure positive
    
    growth_data <- rbind(growth_data, data.frame(
      nutrient = nutrient,
      growth_rate = replicates,
      replicate = 1:6
    ))
  }
  
  # Fit the model
  # Here tryCatch lets me run many models and if one fails it logs it and moves on instead of crashing the loop which is nice.
  tryCatch({ 
    fit <- brm(
      formula = monod_formula,
      data = growth_data,
      prior = priors,
      family = gaussian(),
      chains = 4,
      iter = 4000,
      warmup = 2000,
      cores = 6,
      control = list(adapt_delta = 0.95),
      refresh = 0,  # suppress iteration printing
      silent = 2    # suppress most messages
    )
    
    # Pull estimates and other values for analysis
    mumax_est <- summary(fit)$fixed[1,1]
    mumax_est_er<-summary(fit)$fixed[1,2]
    mumax_ci_lo <-summary(fit)$fixed[1,3]
    mumax_ci_up <-summary(fit)$fixed[1,4]
    mumax_rhat <-summary(fit)$fixed[1,5]
    mumax_bulk_ess<-summary(fit)$fixed[1,6]
    mumax_tail_ess<-summary(fit)$fixed[1,7]
    K_est <- summary(fit)$fixed[2,1]
    K_est_er<-summary(fit)$fixed[2,2]
    K_ci_lo <-summary(fit)$fixed[2,3]
    K_ci_up <-summary(fit)$fixed[2,4]
    K_rhat <-summary(fit)$fixed[2,5]
    K_bulk_ess<-summary(fit)$fixed[2,6]
    K_tail_ess<-summary(fit)$fixed[2,7]
    sigma<-summary(fit)$spec_pars[1,1]
    
    # Store results
    results <- rbind(results, data.frame(
      inoculum = current_inoculum,
      sd = current_sd,
      mumax_estimate = mumax_est,
      mumax_estimate_error = mumax_est_er,
      mumax_lower = mumax_ci_lo,
      mumax_upper = mumax_ci_up,
      mumax_bulk_ess=mumax_bulk_ess,
      mumax_tail_ess=mumax_tail_ess,
      mumax_rhat =mumax_rhat,
      K_estimate = K_est,
      K_estimate_error = K_est_er,
      K_lower =K_ci_lo,
      K_upper = K_ci_up,
      K_bulk_ess=K_bulk_ess,
      K_tail_ess=K_tail_ess,
      K_rhat =K_rhat,
      sigma=sigma
    ))
    
    cat("mumax est:", round(mumax_est, 3), 
        "CI: [", round(mumax_ci_lo, 3), ",", round(mumax_ci_up, 3), "]\n")
    cat("K est:", round(K_est, 3), 
        "CI: [", round(K_ci_lo, 3), ",", round(K_ci_up, 3), "]\n")
    end_time<-Sys.time()
    cat("Model", i, "Runtime", end_time-start_time)
    
    
  }, error = function(e) {
    cat("Fit failed:", e$message, "\n")
  })
}


end_time_total<-Sys.time()

#Total model list runtime
print(end_time_total-start_time_total)

# View all results
print(results)

# Save results
write.csv(results, "monod_fitting_results.csv", row.names = FALSE)

# Plot mu_max estimates
p1 <- ggplot(results, aes(x = factor(inoculum), y = mumax_estimate, color = factor(sd))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mumax_lower, ymax = mumax_upper), width = 0.2) +
  labs(title = "mu_max Estimates Across Conditions",
       x = "Inoculum Size", y = "mu_max", color = "Std Dev") +
  theme_minimal()

# Plot K_s estimates
p2 <- ggplot(results, aes(x = factor(inoculum), y = K_estimate, color = factor(sd))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = K_lower, ymax = K_upper), width = 0.2) +
  labs(title = "K_s Estimates Across Conditions",
       x = "Inoculum Size", y = "K_s", color = "Std Dev") +
  geom_abline(intercept =0.0101,slope=0,aes(colour="red"))+
  theme_minimal()

print(p1)
print(p2)
```

Quantify Bias

```{r}
results2<-mutate(results, K_bias = abs(results$K_estimate-0.0101))
results2<-mutate(results2, mumax_bias = abs(results$mumax_estimate-1))
```

## Plot Bias Contour (first attempt)

```{r}
ggplot(results2,aes(x=log2(inoculum),y=sd,fill=log2(K_bias)))+
  geom_tile()

ggplot(results2,aes(x=log2(inoculum),y=sd,fill=mumax_bias))+
  geom_tile()
```

```{r}
ggplot(results2,aes(x=log2(inoculum),y=sd,z=mumax_bias))+
  geom_contour_filled()

ggplot(results2,aes(x=log2(inoculum),y=sd,z=log2(K_bias)))+
  geom_contour_filled()

results3<-results2
```

Plot initial contours

```{r}
ggplot(results3,aes(x=log2(inoculum),y=sd,z=log(mumax_bias)))+
  geom_contour_filled(bins=10)+  
  theme_classic()+
  theme(legend.position = "right",
        legend.key.size = unit(0.35, "cm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size =9))


ggplot(results3,aes(x=log2(inoculum),y=sd,z=log2(K_bias)))+
  geom_contour_filled()+
  theme_classic()
```

hmmm , these look bad, maybe there are ways to smooth them?

## Plot Bias Contour (second attempt)

Attempt to interpolate data to get smoother contour plots (not used in final analyses)

```{r}
library(akima)

# Interpolate to create a finer grid
interp_data <- interp(
  x = log2(results3$inoculum),
  y = results3$sd,
  z = log2(results3$K_bias),
  xo = seq(min(log2(results3$inoculum)), max(log2(results3$inoculum)), length = 200),
  yo = seq(min(results3$sd), max(results3$sd), length = 200),
  linear = F,  
  extrap = T
)

interp_df <- expand.grid(x = interp_data$x, y = interp_data$y)
interp_df$z <- as.vector(interp_data$z)
interp_df <- na.omit(interp_df)
```

```{r}
K_plot<-ggplot(interp_df, aes(x = x, y = y, z = z)) +
  geom_contour_filled(bins = 9) +
  labs(x = "Inoculum Size (log2)", 
       y = "sd",
       fill = "log2") +
  ggtitle('K Bias')+
  theme_classic()+
  theme(legend.position = "right",
        legend.key.size = unit(0.35, "cm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size =9))
```

```{r}
mumax_plot<-ggplot(results3,aes(x=log2(inoculum),y=sd,z=log2(mumax_bias)))+
  geom_contour_filled(bins=8)+  
  labs(x = "Inoculum Size (log2)", 
       y = "sd",
       fill = "log2") +
  ggtitle('MuMax Bias')+
  theme_classic()+
  theme(legend.position = "right",
        legend.key.size = unit(0.35, "cm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size =9))
```

```{r}
K_plot
mumax_plot
```

## Plot Bias Contour (third attempt) \[BEST\]

Fit GAMs to the bias data instead! Much more statistically rigorous!

```{r}
gam_model_K <- gam(log2(K_bias) ~ s(log2(inoculum), sd, bs = "tp",k=20), 
                   data = results3)
summary(gam_model_K)
gam_model_mumax <- gam(log2(mumax_bias) ~ s(log2(inoculum), sd, bs = "tp",k=20), 
                       data = results3)
summary(gam_model_mumax)

# Create prediction grid
pred_grid <- expand.grid(
  inoculum = seq(min(results3$inoculum), max(results3$inoculum), length = 200),
  sd = seq(min(results3$sd), max(results3$sd), length = 200)
)

pred_grid$predicted_K <- predict(gam_model_K, newdata = pred_grid)
pred_grid$predicted_mumax <- predict(gam_model_mumax, newdata = pred_grid)

# Plot the bias predicted values from GAMs
ggplot(pred_grid, aes(x = log2(inoculum), y = sd, z = predicted_K)) +
  geom_contour_filled(bins = 20) +
  geom_point(data = results3, aes(x = log2(inoculum), y = sd,color=log2(K_bias)), 
             inherit.aes = FALSE, size = 2) +
  scale_color_viridis()+
  ggtitle('K Bias')+
  labs(x = "log2[inoculum size] (cells/ul)", y = "SD", fill = "log2(K_bias)") +
  theme_classic()

ggplot(pred_grid, aes(x = log2(inoculum), y = sd, z = predicted_mumax)) +
  geom_contour_filled(bins = 20) +
  geom_point(data = results3, aes(x = log2(inoculum), y = sd,color=log2(mumax_bias)), 
             inherit.aes = FALSE, size = 2) + 
  scale_color_viridis()+
  ggtitle('mumax Bias')+
  labs(x = "log2[inoculum size] (cells/ul)", y = "SD", fill = "mumax_bias") +
  theme_classic()
```

```{r}
gam.check(gam_model_K)
gam.check(gam_model_K_nolog)

summary(gam_model_K)
```

```{r}
gam.check(gam_model_mumax)
gam.check((gam_model_mumax_nolog))


summary(gam_model_mumax)
```

```{r}
gam_model_mumax_nolog <- gam(mumax_bias ~ s(log2(inoculum), sd, bs = "tp",k=20), 
                             data = results3)

gam.check(gam_model_mumax_nolog)
summary(gam_model_mumax_nolog)
```

```{r}
# Compare your K_bias vs mumax_bias models
AIC(gam_model_mumax_nolog, gam_model_mumax)
BIC(gam_model_mumax_nolog, gam_model_mumax)

# Or create comparison table
comparison <- data.frame(
  response = c("mumax_nolog_bias", "mumax_bias"),
  AIC = c(AIC(gam_model_mumax_nolog), AIC(gam_model_mumax)),
  R_sq = c(summary(gam_model_mumax_nolog)$r.sq, summary(gam_model_mumax)$r.sq),
  edf = c(sum(gam_model_mumax_nolog$edf), sum(gam_model_mumax$edf))
)

comparison
```

```{r}
# Create prediction grid
pred_grid <- expand.grid(
  inoculum = seq(min(results3$inoculum), max(results3$inoculum), length = 200),
  sd = seq(min(results3$sd), max(results3$sd), length = 200)
)

pred_grid$predicted_mumax <- predict(gam_model_mumax_nolog, newdata = pred_grid)

# Plot the bias predicted values from GAMs


ggplot(pred_grid, aes(x = log2(inoculum), y = sd, z = predicted_mumax)) +
  geom_contour_filled(bins = 20) +
  geom_point(data = results3, aes(x = log2(inoculum), y = sd,color=mumax_bias), 
             inherit.aes = FALSE, size = 2) + 
  scale_color_viridis()+
  ggtitle('mumax Bias')+
  labs(x = "log2[inoculum size] (cells/ul)", y = "SD", fill = "mumax_bias") +
  theme_classic()
```

```{r}
gam_model_K_nolog <- gam(K_bias ~ s(log2(inoculum), sd, bs = "tp",k=20), 
                         data = results3)
pred_grid <- expand.grid(
  inoculum = seq(min(results3$inoculum), max(results3$inoculum), length = 200),
  sd = seq(min(results3$sd), max(results3$sd), length = 200)
)

pred_grid$predicted_mumax <- predict(gam_model_K_nolog, newdata = pred_grid)

# Plot the bias predicted values from GAMs


ggplot(pred_grid, aes(x = log2(inoculum), y = sd, z = predicted_mumax)) +
  geom_contour_filled(bins = 20) +
  geom_point(data = results3, aes(x = log2(inoculum), y = sd,color=K_bias), 
             inherit.aes = FALSE, size = 2) + 
  scale_color_viridis()+
  ggtitle('K Bias')+
  labs(x = "log2[inoculum size] (cells/ul)", y = "SD", fill = "K_bias") +
  theme_classic()
```

```{r}
# Compare your K_bias vs mumax_bias models
AIC(gam_model_K_nolog, gam_model_K)
BIC(gam_model_K_nolog, gam_model_K)

# Or create comparison table
comparison <- data.frame(
  response = c("K_nolog_bias", "K_bias"),
  AIC = c(AIC(gam_model_K_nolog), AIC(gam_model_K)),
  BIC = c(BIC(gam_model_K_nolog), BIC(gam_model_K)),
  R_sq = c(summary(gam_model_K_nolog)$r.sq, summary(gam_model_K)$r.sq),
  edf = c(sum(gam_model_K_nolog$edf), sum(gam_model_K$edf))
)

comparison
```
